---
title: "a3_task4_daija_odom"
author: "Daija Odom"
date: "10/28/2021"
output: 
  html_document: 
    theme: cosmo
    toc: yes
---
# Overview

This code wrangles, visualizes, and analyzes data for mean size of purple urchins at the Naples Reef long term kelp removal experiment sites in 2010. 

**Data citation:** Reed, D, R. Miller. 2021. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 21. Environmental Data Initiative. https://doi.org/10.6073/pasta/fd564dddfe7b77fe9e4bd8417f166057.

# Setup

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)

library(janitor)
library(here)
library(lubridate)
library(kableExtra)

```

# Read in the data

```{r}
urchins <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names()
```

# Data wrangling: wrangle data to only include purple urchins observations at Naples Reef in 2010 

```{r}
urchins_naples <- urchins %>% 
  filter(common_name == "Purple Urchin", # Filter only purple urchins 
         site == "NAPL") %>% # Filter by site of Naples Reef 
  mutate(date = lubridate::mdy(date)) %>% # Converted class to date
  mutate(year = lubridate::year(date)) %>% # Added a new column for year
  filter(year == 2010) # Kept only observation from 2010 in year column
```

# Data visualization: create a histogram and Q-Q plot of size distribution for purple urchins at Naples Reef in 2010 separated by treatment

```{r}
ggplot(data = urchins_naples, aes(x = size, fill = treatment)) +
  geom_histogram(bins = 10) + # Reduced the number of bins from default
  facet_wrap(~ treatment) + # Split column into two by treatment
  labs( x = "Size of Urchins (cm)",
        y = "Count",
        title = "Distribution of Size of Purple Urchins",
        fill = "Treatment") # Change the legend title using `fill =` for histogram

```


```{r}
urchins_qq <- ggplot(data = urchins_naples) + 
  geom_qq(aes(sample = size)) + # Create a Q-Q plot
  geom_qq_line(aes(sample = size)) +
  facet_wrap( ~ treatment) + # Divide column by annual and control treatment
  labs(x = "Theoretical Size (cm)",
       y = "Sample Size (cm)",
       title = "Distribution of Size of Purple Urchins")

urchins_qq # Call on histogram
  
```

**Main takeaways:** The two exploratory graphs show a normal distribution between purple urchins in the annual and controlled sites. Based on the line in the Q-Q plot urchins in the controlled site appear to be more normally distributed based on the skew towards 3 cm. 

# Statistics Table: finalized summary table of mean, standard deviation, and sample size for purple urchins in 2010

```{r}
urchins_naples_2010 <- urchins_naples %>% # Created subset of data to only include treatment and measures of central tendency
  group_by(treatment) %>% 
  summarize("Mean Diameter (cm)" = round(mean(size),1), # Collapsed original rows into single row per group
            "Standard Deviation" = round(sd(size),1), # Rounded to 1 decimal place
            "Sample Size" = n()) %>% 
  rename("Treatment" = treatment) # Renamed treatment header to look cleaner

 knitr::kable(head(urchins_naples_2010),
              caption = "Table 1. Mean, standard deviation, and sample size of purple urchins") %>% 
   kable_styling(full_width = FALSE) # Fix table column headers
 

```

